Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-27 icfp.tex
erased trivial cases of subst lemmas

\author{Rodrigo Mesquita}
%\affiliation{
  %\department{Departamento de Inform\'{a}tica}
  %\institution{FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  %\country{Portugal}}
\affiliation{
   \institution{Well-Typed LLP}
  %\city{City}
  %\country{Something}
 }
 \email{rodrigo@well-typed.com}
\author{Bernardo Toninho}
\orcid{0000-0002-0746-7514}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{NOVA-LINCS, FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\email{btoninho@fct.unl.pt}
%% TODO: AFFILIATION, EMAIL
%% TODO: CCS CONCEPTS and KEYWORDS
%% TODO: ACM RIGHTS notice...
% CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf

@@

The instrumented semantics is presented in
Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
contexts for expressions in $\Sigma$
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e : \tau$ which are expressions in which
bindings in $\Theta$ may also be used. This additional data is needed to
inductively show that the overall evaluation state is well-typed (and
linearity preserving). The rules mostly mirror those of the natural
semantics, with the exception of the linear variable rule which erases
the binding from the execution environment. Both semantics define
evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
are evaluation states. We also rely on a notion
of partial derivations and partial evaluation (see
Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
to state and reason about a progress property.\todo{Na semantica O omega e 1 ainda estão separados, é assim?}

@@

    \item Programs rejected due to the lack of ``linearity coercions'' in
        $\lambda^\pi_\Delta$. So-called \emph{multiplicity coercions} extending
        Linear Core could make these programs be understood as linear
        (discussed in Section~\ref{sec:future-work}).

@@

\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}.}
%
% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}
%
% Haskell, contrary to most programming languages with linear types, has existed
% for 31 years of its life \emph{without} linear types. As such, the introduction
% of linear types to Haskell comes with added challenges that do not exist in
% languages that were designed with linear types from the start:
% %
% \begin{itemize}
%     \item Backwards compatibility. The addition of linear types shouldn't break
%         all existing Haskell code.
%     \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
%         its non-linearly-typed counterpart should fit in together and it must be
%         possible to define functions readily usable by both sides
%         simultaneously.
%     \item Future-proofing. Haskell, despite being an
%         industrial-strength language, is also a petri-dish for experimentation
%         and innovation in the field of programming languages. Therefore, Linear
%         Haskell takes care to accommodate possible future features, in
%         particular, its design is forwards compatible with affine and dependent
%         types.
% \end{itemize}
% %
Linear Haskell~\cite{cite:linearhaskell} augments the Haskell surface language
with linear types, but it is not concerned with extending GHC's intermediate
language(s), which presents its own challenges. 
%
However, the implementation of Linear Haskell in GHC does modify and extend Core
with linearity/multiplicity annotations. Core's type system is unable to type
semantic linearity of programs in the sense elaborated in our work,
which results in Core rejecting most linear programs resulting from optimising
transformations that leverage the non-strict semantics of Core.
%
Our work overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding linearity semantically in the presence of
laziness, and by proving linearity-preserving multiple Core-to-Core
optimisations GHC.
Linear Core can also be seen as a system that validates the programs written in
Linear Haskell and are compiled by GHC, by guaranteeing (through typing) that
linear resources are still used exactly once throughout the optimising
transformations.\todo{add part about operational semantics and proof being similar to them}

@@

Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation
environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
$\Theta\downarrow^1_e$ refers to the expressions bound to linear variables in
the runtime environment and $\Theta\downarrow^1_\sigma$ refers to the types of
those expressions.

 \begin{definition}[Environment Expansion]
   We rely on an expansion of a runtime environment $\Theta$ for the evaluation of a term $e$ into a (typed) term, written $[\Theta \mid e]$, defined inductively as follows:
   \[
     \begin{array}{lcl}
       \left[ \cdot \mid e \right] & \triangleq & e\\
       \left[ \Theta , \xl = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda \xl . e)~e']\\
       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & [\Theta \mid \llet{x:_\Delta \sigma = e'}{e}]
    \end{array}
   \]
 \end{definition}

@@

 \begin{lemma}~\label{lem:unwrapenv}
   Iff $\G ; \D \vdash [ \Theta \mid e] : \sigma$
     then $\G , \Theta\uparrow^{\omega} ; \D , \Theta\uparrow^1 \vdash  (e, \Theta\downarrow^1_e) : \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$
   \end{lemma}
\begin{proof}
By structural induction on the environment expansion definition.
\begin{description}
%\item[Case:] $\left[\cdot \mid e\right] \triangleq e$ Trivial
\item[Case:] $\left[\Theta, \xl = y \mid e\right] \triangleq \left[\Theta \mid (\lambda \xl. e)~y\right]$
\begin{tabbing}
    (1) $\G; \D \vdash \left[ \Theta, \xl = y \mid e \right] : \phi$\\
    (2) $\G; \D \vdash \left[ \Theta \mid (\lambda \xl. e)~y \right] : \phi$\` by def.\\
    (3) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash ((\lambda \xl. e)~y, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
    (4) $Let~\D_1,\D_2,\D_3 = \D, \Theta\uparrow^1 $\\
    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
    (6) $\G, \Theta\uparrow^\omega; \D_2 \vdash y : \s$\` by inv\\
    (7) $\G, \Theta\uparrow^\omega; \D_1 \vdash (\lambda \xl. e) : \s \to \phi $\` by inv\\
    (8) $\G, \Theta\uparrow^\omega; \D_1, \xl \vdash e : \phi $\` by inv\\
    (9) $\G, \Theta\uparrow^\omega; \D_1, \D_2, \xl \vdash (e, y) : \phi \otimes \sigma$\` by pair\\
    (10) $\G, \Theta\uparrow^\omega; \D, \xl \vdash (e, y, \Theta\downarrow^1_e) : \phi \otimes \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
\end{tabbing}
\item[Case:] $\left[\Theta, \xD = e' \mid e\right] \triangleq \left[\Theta \mid \llet{\xD = e'}{e}\right]$
\begin{tabbing}
    (1) $\G; \D' \vdash \left[ \Theta, \xD = y \mid e \right] : \phi$\\
    (2) $\G; \D' \vdash \left[ \Theta \mid \llet{\xD = e'}{e} \right] : \phi$\` by def.\\
    (3) $\G, \Theta\uparrow^\omega; \D', \Theta\uparrow^1 \vdash (\llet{\xD = e'}{e}, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
    (4) $Let~\D,\D_2,\D_3 = \D', \Theta\uparrow^1 $\\
    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
    (6) $\G, \Theta\uparrow^\omega; \D \vdash e' : \s$\` by inv on let\\
    (7) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2 \vdash e : \phi $\` by inv on let\\
    (8) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2,\D_3 \vdash (e, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
\end{tabbing}
\end{description}
\end{proof}

@@

\begin{proof}
By structural induction on the instrumented operational semantics reductions.
\begin{description}
    \item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
\begin{tabbing}
    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\} : \tau, \Sigma$\\
    (2) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\}, \Theta\downarrow^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta\downarrow^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\to e'\}} : \tau$\\
    Subcase $e$ is in WHNF (thus $= K~\ov{x_i}$)\\
    (5) $\G'; \D_a \Vdash K~\ov{x_i} : \s \gtrdot \ov{\D_i}$\` by subcase\\
    (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
    Subsubcase $\vdash_{alt}$ $Alt_{NWHNF}$\\
    (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
    (8) $\Theta' = \Theta$\` in this subcase because $(\Theta \mid K~\ov{x_i})\Downarrow(\Theta \mid K~\ov{x_i})$\\
    (9) $\G'; \ov{\D_i} \vdash K~\ov{x_i} : \s$\` by inv (5)\\
    (10) $\ov{\G'; \cdot \vdash x_i : \s_i}$\` by inv (9) for unrestricted $x$s\\
    (11) $\ov{\G'; \D_j \vdash x_j : \s_j}$\` by inv (9) for linear fields $x$s\\
    (12) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}] : \tau}$\` by $\D$-subst and $\omega$-subst lemmas (7,10,11)\\
    (13) $\G'; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\` by $\D$-subst lemma and $\ov{\D_i} = \D_a$ (9,12)\\
    (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by reconstructing\\
    (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
    Subsubcase $\vdash_{alt}$ $Alt_{0}$\\
    (7) $(\G', z{:}_{\D_a}\s)[\cdot/\D_a]_z, \ov{y_i{:}_\omega\sigma_i}; (\D_a, \D_b)[\cdot / \D_a] \vdash e' : \tau$\` by inv subcase\\
    (8) $\G', z{:}_{\cdot}, \ov{y_i{:}_\omega\sigma_i}; \D_b \vdash e' : \tau$\` by $z$ cannot occur in $\G'$\\
    (9) Since $\G';\D_a \Vdash K~\ov{x_i}$, in this subcase $\D_a = \cdot$\\
    (10) $\G', \z{:}_\cdot\sigma, \ov{y{:}_\omega\sigma}; \D_b \vdash e' : \tau$\` by (9)\\
    (11) $\G', \z{:}_\cdot\sigma; \D_b \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst (10)\\
    (12) $\G'; \D_b \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst\\
    (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
    (14) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
    Subcase $e$ is not in WHNF\\
    (5) $\G'; \D_a \vdash e : \s$\` by inv on case\\
    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
\end{tabbing}
\end{description}
\end{proof}

>>>>

\author{Rodrigo Mesquita}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\affiliation{
   \institution{Well-Typed LLP}
  %\city{City}
  %\country{Something}
 }
 \email{xxx@yyy.com}
\author{Bernardo Toninho}
\orcid{0000-0002-0746-7514}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{NOVA-LINCS, FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\email{btoninho@fct.unl.pt}
%% TODO: AFFILIATION, EMAIL
%% TODO: CCS CONCEPTS and KEYWORDS
%% TODO: ACM RIGHTS notice...
% CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf

@@

The instrumented semantics is presented in
Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
contexts for expressions in $\Sigma$
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e : \tau$ which are expressions in which
bindings in $\Theta$ may also be used. This additional data is needed to
inductively show that the overall evaluation state is well-typed (and
linearity preserving). The rules mostly mirror those of the natural
semantics, with the exception of the linear variable rule which erases
the binding from the execution environment. Both semantics define
evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
are evaluation states. We also rely on a notion
of partial derivations and partial evaluation (see
Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
to state and reason about a progress property.

@@

    \item Programs rejected due to the lack of ``linearity coercions'' in
        $\lambda^\pi_\Delta$. Extending Linear Core with \emph{multiplicity coercions} 
        could make these programs be understood as linear
        (discussed in Section~\ref{sec:future-work}).

@@

\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}.}
%
% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}
%
% Haskell, contrary to most programming languages with linear types, has existed
% for 31 years of its life \emph{without} linear types. As such, the introduction
% of linear types to Haskell comes with added challenges that do not exist in
% languages that were designed with linear types from the start:
% %
% \begin{itemize}
%     \item Backwards compatibility. The addition of linear types shouldn't break
%         all existing Haskell code.
%     \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
%         its non-linearly-typed counterpart should fit in together and it must be
%         possible to define functions readily usable by both sides
%         simultaneously.
%     \item Future-proofing. Haskell, despite being an
%         industrial-strength language, is also a petri-dish for experimentation
%         and innovation in the field of programming languages. Therefore, Linear
%         Haskell takes care to accommodate possible future features, in
%         particular, its design is forwards compatible with affine and dependent
%         types.
% \end{itemize}
% %
Linear Haskell~\cite{cite:linearhaskell} augments the Haskell surface language
with linear types, but it is not concerned with extending GHC's intermediate
language(s), which presents its own challenges. 
%
However, the implementation of Linear Haskell in GHC does modify and extend Core
with linearity/multiplicity annotations. Core's type system is unable to type
semantic linearity of programs in the sense elaborated in our work,
which results in Core rejecting most linear programs resulting from optimising
transformations that leverage the non-strict semantics of Core.
%
Our work overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding linearity semantically in the presence of
laziness, and by showing that multiple Core-to-Core
optimisations employed by GHC are linearity preserving, using a
linearity-aware semantics that is essentially identical to that of~\cite{cite:linearhaskell}.
Linear Core can also be seen as a system that validates the programs written in
Linear Haskell and are compiled by GHC, by guaranteeing (through typing) that
linear resources are still used exactly once throughout the optimising
transformations.

@@

Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.

 \begin{definition}[Environment Expansion]
   We rely on an expansion of a runtime environment $\Theta$ for the evaluation of a term $e$ into a (typed) term, written $[\Theta \mid e]$, defined inductively as follows:
   \[
     \begin{array}{lcl}
       \left[ \cdot \mid e \right] & \triangleq & e\\
       \left[ \Theta , x:_1\sigma = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda x_1:\sigma . e)~e']\\
       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & [\Theta \mid \llet{x:_\Delta \sigma = e'}{e}]
    \end{array}
   \]
 \end{definition}

@@

 \begin{lemma}
   Iff $\G ; \D \vdash [ \Theta \mid e] : \sigma$
     then $\G , \Theta\uparrow^{\omega} ; \D , \Theta\uparrow^1 \vdash  (e, \Theta\downarrow^1_e) : \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$
   \end{lemma}

@@
